Imputation of Incomplete Multilevel Data with \pkg{mice}

### 2022-12-12 Manuscript/Imputation_of_Incomplete_Multilevel_Data.Rmd
tables

---
documentclass: jss
author:
  - name: Hanne I. Oberman
    affiliation: Utrecht University
    address: |
      | Padualaan 14
      | 3584 CH Utrecht
    email: \email{h.i.oberman@uu.nl}
    url: https://hanneoberman.github.io/
  - name: "Johanna Munoz Avila"
    affiliation: Julius Center for Health Sciences and Primary Care, University Medical Center Utrecht, Utrecht University, Utrecht, The Netherlands 
  - name: Thomas P. A. Debray
    affiliation: Julius Center for Health Sciences and Primary Care, University Medical Center Utrecht, Utrecht University, Utrecht, The Netherlands 
  - name: Gerko Vink
    affiliation: Utrecht University 
  - name: Valentijn M. T. de Jong
    affiliation: Julius Center for Health Sciences and Primary Care, University Medical Center Utrecht, Utrecht University, Utrecht, The Netherlands \AND Data Analytics and Methods Task Force, European Medicines Agency, Amsterdam, The Netherlands

title:
  formatted: "Imputation of Incomplete Multilevel Data with \\pkg{mice}"
  plain:     "Imputation of Incomplete Multilevel Data with mice"
  short:     "Multilevel \\pkg{mice}"
abstract: >
  This tutorial illustrates the imputation of incomplete multilevel data with the \proglang{R} packackage \pkg{mice}. Footnotes in the current version show work in progress/under construction. The last section is not part of the manuscript, but purely for reminders. See also all of the TODOs that need to be worked out. We aim to submit at JSS, so there is no word count limit ("There is no page limit, nor a limit on the number of figures or tables"). [Just adding some text to get a better guess of what the actura abstract will look like: Lorem ipsum dolor sit amet, consectetur adipiscing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua. Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat. Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur. Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.]
keywords:
  # at least one keyword must be supplied
  formatted: [missing data, multilevel, clustering, "\\pkg{mice}", "\\proglang{R}"]
  plain:     [missing data, multilevel, clustering, mice, R]
preamble: >
  \usepackage{amsmath}
header-includes: 
 - \usepackage{graphicx} 
 - \usepackage{mathtools}
 - \usepackage{ulem}
output: 
    rticles::jss_article:
      keep_tex: yes
      number_sections: yes
    # word_document: default
bibliography: ../References/multilevelmice.bib
editor_options: 
  chunk_output_type: inline
---

@@

Many datasets include individuals from multiple settings, geographic regions, or even different studies. In the simplest case, individuals (e.g., students) are nested within so-called clusters (e.g., school classes). More complex clustered structures may occur when there are multiple hierarchical levels (e.g., patients within hospitals within regions or countries), or when the clustering is non-nested (e.g., electronic health record data from diverse settings and populations within large databases). In general, individuals from the same cluster tend to be more similar than individuals from other clusters. In statistical terms, this implies that observations from the same cluster are correlated. If this correlation is left unaddressed, estimates of *p* values, confidence intervals even model parameters are prone to bias  [@loca01]. [TODO: make a link to imputation methods, which require adequate handling and propagation of variance; we are not recommending the adoption of multilevel models for data analysis here, but rather for imputation. VMTdJ: "I would only make that link after introducing missing data, and missing data is not mentioned yet, so I would do that in the next section"] Statistical methods for clustered data typically adopt hierarchical models that explicitly describe the grouping of observations. These models are also known as 'multilevel models', 'hierarchical models', 'mixed effect models', 'random effect models', and in the context of time-to-event data as 'frailty models'. Table \ref{tab:clus} provides an overview of some key concepts in multilevel modeling.

\begin{table}[tb]
\caption{Concepts in multilevel methods}
\label{tab:clus}
\centering
\begin{tabular}{ll}
\hline
\textbf{Concept} & \textbf{Details}   \\
\hline
ICC                 & The variability due to clustering is often measured by means of the \\
                    & intraclass coefficient (ICC). The ICC can be seen as the percentage \\
                    & of variance that can be attributed to the cluster-level, where a high \\
                    & ICC would indicate that a lot of variability is due to the cluster \\
                    & structure. \\
Random effect       & Multilevel models typically accommodate for variability by including \\
                    & a separate group mean for each cluster. In addition to random \\
                    & intercepts, multilevel models can also include random coefficients \\
                    & and heterogeneous residual error variances across clusters [see e.g. \\
                    & @gelm06, @hox17 and @jong21]. [TODO: add stratification.] \\
\hline
\end{tabular}
\end{table}

@@

Imputation of missing data requires consideration of the mechanism behind the missingness. Rubin proposed to distinguish between data that are missing completely at random (MCAR), data that are missing at random (MAR) and data that are missing not at random (MNAR; see Table \ref{tab:miss}). For each of these three missingness generating mechanisms, different imputation strategies are warranted @yuce08 and @hox15. We here consider the general case that data are MAR, and expand on certain MNAR situations.

@@

The \proglang{R} package \pkg{mice} has become the de-facto standard for imputation by chained equations, which iteratively solves the missingness on a variable-by-variable basis. \pkg{mice} is known to yield valid inferences under many different missing data circumstances [@buur18]. However, commonly used imputation methods were not designed for use in clustered data and usually generate observations that are independent whereas multilevel data are dependent.  For this reason, we discuss how the \proglang{R} package \pkg{mice} can be used to impute multilevel data.

@@

This papers serves as a tutorial for imputing incomplete multilevel data with \pkg{mice} in \proglang{R}. We provide practical guidelines and code snippets for different missing data situations, including non-ignorable mechanisms. For reasons of brevity, we focus on multilevel imputation by chained equations with \pkg{mice} exclusively; other imputation methods and packages (e.g., \pkg{jomo} and \pkg{mdmb}) are outside the scope of this tutorial. [TODO: change this sentence because we do not assume familiarity: "Assumed knowledge includes basic familiarity with multilevel imputation [see e.g. @audi18 and @grun18] and the \pkg{lme4} notation for multilevel models (see Table \ref{tab:mod})"].

\begin{table}[tb]
\caption{Notation}
\label{tab:mod}
\centering
\begin{tabular}{ll}
\hline
\textbf{Concept} & \textbf{Details}   \\
\hline
& [TODO: explain \pkg{lme4} notation here] \\
\hline
\end{tabular}
\end{table}

@@

This tutorial is dedicated to readers who are unfamiliar with multiple imputation. More experienced readers can skip the introduction (case study 1) and directly head to practical applications of multilevel imputation under MAR conditions (case study IMPACT) or  under MNAR conditions (case study HIV).

@@

Impute the data:
```{r}
imp <- mice( hivdata, # dataset with missing values
                        m = 10,   # number of imputations
                        maxit = 1,
                        seed = 1234, #seed attached to the dataID
                        meth = meth, #imputation method vector
                        pred = pred, #imputation predictors matrix
                        print = T,
                        meta_method="reml",
                        pmm=FALSE)
```

>>>>

---
documentclass: jss
author:
  - name: Hanne Oberman
    affiliation: Utrecht University
    address: |
      | Padualaan 14
      | 3584 CH Utrecht
    email: \email{h.i.oberman@uu.nl}
    url: https://hanneoberman.github.io/
  - name: "Johanna Munoz"
    affiliation: University Medical Center Utrecht \AND
  - name: Valentijn de Jong
    affiliation: University Medical Center Utrecht 
  - name: Gerko Vink
    affiliation: Utrecht University \AND
  - name: Thomas Debray
    affiliation: University Medical Center Utrecht 
title:
  formatted: "Imputation of Incomplete Multilevel Data with \\pkg{mice}"
  plain:     "Imputation of Incomplete Multilevel Data with mice"
  short:     "Multilevel \\pkg{mice}"
abstract: >
  This tutorial illustrates the imputation of incomplete multilevel data with the \proglang{R} packackage \pkg{mice}. Footnotes in the current version show work in progress/under construction. The last section is not part of the manuscript, but purely for reminders. See also all of the TODOs that need to be worked out. We aim to submit at JSS, so there is no word count limit ("There is no page limit, nor a limit on the number of figures or tables"). [Just adding some text to get a better guess of what the actura abstract will look like: Lorem ipsum dolor sit amet, consectetur adipiscing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua. Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat. Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur. Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.]
keywords:
  # at least one keyword must be supplied
  formatted: [missing data, multilevel, clustering, "\\pkg{mice}", "\\proglang{R}"]
  plain:     [missing data, multilevel, clustering, mice, R]
preamble: >
  \usepackage{amsmath}
header-includes: 
 - \usepackage{graphicx} 
 - \usepackage{mathtools}
 - \usepackage{ulem}
output: 
    rticles::jss_article:
      keep_tex: yes
      number_sections: yes
    # word_document: default
bibliography: ../References/multilevelmice.bib
editor_options: 
  chunk_output_type: inline
---

@@

Many datasets include individuals from multiple settings, geographic regions, or even different studies. In the simplest case, individuals (e.g., students) are nested within so-called clusters (e.g., school classes). More complex clustered structures may occur when there are multiple hierarchical levels (e.g., patients within hospitals within regions or countries), or when the clustering is non-nested (e.g., electronic health record data from diverse settings and populations within large databases). In general, individuals from the same cluster tend to be more similar than individuals from other clusters. In statistical terms, this implies that observations from the same cluster are correlated. If this correlation is left unaddressed, estimates of *p* values, confidence intervals even model parameters are prone to bias  [@loca01]. [TODO: make a link to imputation methods, which require adequate handling and propagation of variance; we are not recommending the adoption of multilevel models for data analysis here, but rather for imputation.] Statistical methods for clustered data typically adopt hierarchical models that explicitly describe the grouping of observations. These models are also know as 'multilevel models', 'hierarchical models', 'mixed effect models' and 'random effect models'. Table \ref{tab:clus} provides an overview of some key concepts in multilevel modeling.

\begin{table}[tb]
\caption{Concepts in multilevel methods}
\label{tab:clus}
\centering
\begin{tabular}{ll}
\hline
\textbf{Concept} & \textbf{Details}   \\
\hline
Sample unit         & Units of the population from which measurements are taken in a sample.\\                           & These should be distinct and not overlapping, e.g., patients, students.\\ 
Hierarchical levels & Data are grouped into clusters at different levels. A three-level \\                               & hierarchical model could occur by collecting observations of math scores\\                         & (lower level) which can be measured over the whole school year for each\\                          & student, and then grouped by student (middle level) and these in turn can\\                        & be aggregated into classrooms (high level).\\
Fixed effect        & Here we assume that the values of an independent variable are fixed, i.e., \\ 
  & the values observed in the study are representative of all values in the \\                        & population we are interested in. In the case of a fixed factor variable, for \\                     & example treatment type, we are only concerned with assessing the difference \\ 
                & in the dependent variable y e.g., blood pressure between treatments A and B.\\                     & There may be additional treatments, but based on the objective of the study,\\                     & which in this case is the equivalence effect of A and B, the other treatment\\                     & types are irrelevant and therefore the treatment type is included as a fixed \\                     & effect with only the A and B levels in the hierarchical model. \\
Random effect       & The values of an independent variable are assumed to be randomly drawn from \\                     & a population with more values. In a random factor, the levels observed in the\\                     & study represent a sample of all possible levels, e.g., hospitals, classrooms.\\                     & For example, in a study to predict the blood pressure of patients on \\  
& admission we might select only certain hospitals that are representative of \\                     & the entire population of hospitals. Here we are not interested in measuring \\ 
& the difference of y between individual hospitals, but rather the variation of\\                     & y between hospitals, to create a model that can be generalized to the entire\\                     & population of hospitals. \\

ICC                 & The variability due to clustering is often measured by means of the \\
                    & intraclass coefficient (ICC). The ICC can be seen as the percentage \\
                    & of variance that can be attributed to the cluster-level, where a high \\
                    & ICC would indicate that a lot of variability is due to the cluster \\
                    & structure. \\
Random effect       & Multilevel models typically accommodate for variability by including \\
                    & a separate group mean for each cluster. In addition to random \\
                    & intercepts, multilevel models can also include random coefficients \\
                    & and heterogeneous residual error variances across clusters [see e.g. \\
                    & @gelm06, @hox17 and @jong21]. [TODO: add stratification.] \\
\hline
\end{tabular}
\end{table}
```

@@

Imputation of missing data requires to consider the mechanism behind the missingness. Rubin proposed to distinguish between data that are missing completely at random (MCAR), data that are missing at random (MAR) and data that are missing not at random (MNAR; see Table \ref{tab:miss}). For each of these three missingness generating mechanisms, different imputation strategies are warranted @yuce08 and @hox15. We here consider the general case that data are MAR, and expand on special MNAR situations.

@@

The \proglang{R} package \pkg{mice} has become the de-facto standard for imputation by chained equations, which iteratively solves the missingness on a variable-by-variable basis. \pkg{mice} is known to yield valid inferences under many different missing data circumstances [@buur18]. However, commonly used imputation methods were not designed for use in clustered data and usually generate observations that are independent. For this reason, we discuss how the \proglang{R} package \pkg{mice} can be used to impute multilevel data.

@@

This papers serves as a tutorial for imputing incomplete multilevel data with \pkg{mice} in \proglang{R}. We provide practical guidelines and code snippets for different missing data situations, including non-ignorable mechanisms. For reasons of brevity, we focus on multilevel imputation by chained equations with \pkg{mice} exclusively; other imputation methods and packages (e.g., \pkg{jomo} and \pkg{mdmb}) are outside the scope of this tutorial. Assumed knowledge includes basic familiarity with multilevel imputation [see e.g. @audi18 and @grun18] and the \pkg{lme4} notation for multilevel models (see Table \ref{tab:mod}).

```{=tex}
\begin{table}[tb]
\caption{Notation}
\label{tab:mod}
\centering
\begin{tabular}{lll}
\hline
\textbf{Formula lme4} & \textbf{Details}   \\
\hline
y$\sim$ x1 + (1 | g1)                                       & Fixed x1 predictor with random intercept  \\
                                                         & varying among g\\
y$\sim$x1*x2+ (1| g1)                                      & Interactions of x1 and x2 only in fixed effect\\
y$\sim$x1*x2+ (x2| g1)                                     & Interactions of x1 and x2 only in fixed effect\\
                                                         & with slope of x2 randomly varying among g1\\
y$\sim$x1*x2+ (x1*x2| g1)                                  & variance-covariance matrix estimated only with \\
                                                         & the variance terms of intercept, slope of x1, \\
                                                         & slope of x2 and interaction x1*x2\\
y$\sim$x1*x2+ (x1| g1)+ (x2| g1)                           & variance-covariance matrix estimated separately,\\
                                                         & i.e, one for intercept and x1 and another for \\
                                                         & intercept and x2\\
y$\sim$x1 + (x1 | g1) or 1 + x1 + (1 + x1 | g1)            & Fixed x1 with correlated random intercept and \\
                                                         & random slope of x\\
y$\sim$x1 + (x1 || g1) or 1 + x1 + (1 | g1) + (0 + x1 | g1) & Fixed x1 with uncorrelated random intercept\\
                                                         & and random slope of x1\\  
y$\sim$(1 | g1) + (1 | g2)                                 & Random intercept varying among g1 and among g2\\
y$\sim$(1 | g1/g2) or ~(1|g1)+(1|g1:g2)                    & Random intercept varying among g1 and g2 
                                                         & withing g1\\

& [TODO: explain \pkg{lme4} notation here
https://datalorax.github.io/equatiomatic/articles/lmer4-lmer.html
https://cran.r-project.org/web/packages/lme4/vignettes/lmer.pdf
] \\
\hline
\end{tabular}
\end{table}
```

@@

This tutorial is dedicated to readers who are unfamiliar with multiple imptuation. More experienced readers can skip the introduction (case study 1) and directly head to practical applications of multilevel imputation under MAR conditions (case study IMPACT) or  under MNAR conditions (case study HIV).

@@

<!-- Impute the data: -->
<!-- ```{r} -->
<!-- imp <- mice( hivdata, # dataset with missing values -->
<!--                         m = 10,   # number of imputations -->
<!--                         maxit = 1, -->
<!--                         seed = 1234, #seed attached to the dataID -->
<!--                         meth = meth, #imputation method vector -->
<!--                         pred = pred, #imputation predictors matrix -->
<!--                         print = T, -->
<!--                         meta_method="reml", -->
<!--                         pmm=FALSE) -->
<!-- ``` -->
